---
title: "Case study 1"
author: "PT"
date: "2024-06-26"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(tidyboot)
library(brms)
library(tidybayes)
library(ggpattern)
library(cspplot)
```

# Case study 1

The following script is for processing and running statistical analyses for case study 1. The script uses a dataset of human responses that was manually annotated. Erroneous answers (e.g., false, given the context) were excluded from the data file and the analyses. Only main trials are analysed.

```{r}
d_clean <- read_csv('../data/human/case_study_1/e1_free_production_human_categorized.csv')
```

Run some descriptive analyses:
```{r}
d_clean %>% count(condition, category)

d_clean %>% group_by(condition) %>%
  mutate(condition_count = n()) %>%
  ungroup() %>%
  group_by(condition, category) %>%
  summarise(response_prop = n() / condition_count) %>% unique()

d_clean_taciturn <- d_clean %>% mutate(is_taciturn = ifelse(category == "taciturn", 1, 0)) 
d_clean_taciturn %>%
  group_by(condition) %>%
  tidyboot_mean(column = is_taciturn)
```

```{r}
d_clean_exhaustive <- d_clean %>% mutate(is_exhaustive = ifelse(category == "fullList", 1, 0)) 
d_clean_exhaustive %>%
  group_by(condition) %>%
  tidyboot_mean(column = is_exhaustive)
```

## Stats

We run two Bayesian logistic regressions, analysing whether there is a credible difference in the rate of 1) exhaustive and 2) taciturn responses in the human data, depending on the condition. 
```{r}
lm_exhaustive <- brm(is_exhaustive ~ condition + (1|submission_id),
                     data = d_clean_exhaustive,
                     family = "bernoulli",
                     chains = 4,
                     iter = 3000)
summary(lm_exhaustive)
```

```{r}
lm_taciturn <- brm(is_taciturn ~ condition + (1|submission_id),
                     data = d_clean_taciturn,
                     family = "bernoulli",
                     chains = 4,
                     iter = 3000)
summary(lm_taciturn)
```
